LabMate: supporting types for MATLAB

### 2024-04-09 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] fixing Section 2

\title{LabMate: supporting types for MATLAB}
\author{Conor McBride  \affiliationmark{1}, Georgi Nakov  \affiliationmark{1}, Fredrik Nordvall Forsberg  \affiliationmark{1}, Alistair Forbes  \affiliationmark{2}, Keith Lines  \affiliationmark{2}}
\affiliation{%
\affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , Glasgow, UK}  \\
\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, Teddington, UK}}

@@

This calculation can be converted into a MATLAB problem with LabMate annotations as follows. Note that for reasons of space only twelve values are listed. A real measurement would contain considerably more values:
\begin{verbatim}
%> A :: [ 12 x 5 ] double
A = [ 1  0  1  0  t1
      1  0  1  0  t2
      1  0  1  0  t3
     -1  0  1  0  t4
     -1  0  1  0  t5
     -1  0  1  0  t6
      0  1  0  1  t7
      0  1  0  1  t8
      0  1  0  1  t9
      0 -1  0  1  t10
      0 -1  0  1  t11
      0 -1  0  1  t12 ]

%> B :: [ 5 x 1 ] double
B = [ a aCal c cCal m ]

C = A * B

%> typeof C
\end{verbatim}
%
\todo{Is computing C from A and B what we want? \\ Keith 05/04/2024 ***No. We are computing B from A and C using a least squares line fit.*** \\ FNF 08/04/2024 Should the above be rewritten to do such a line fit, then?}
The lines starting with `\texttt{\%>}' are interpreted as ordinary comments by MATLAB, but as \remph{input directives} by LabMate. LabMate acts by reading the source code and outputting it again, but possibly with some additional \remph{responses} to the directives added.

Did you notice the error in the above code? The matrix \texttt{B} should be of size $5 \times 1$, but is erroneously of size $1 \times 5$ instead. LabMate detects this error and produces the following as output, where we have suppressed the same definition of \texttt{A} as above:
\begin{verbatim}
%> A :: [ 12 x 5 ] double
A = ...

%> B :: [ 5 x 1 ] double
B = [ a aCal c cCal m ]

C = A * B

%> typeof C
%< the type of C is quite a puzzle;
%<   unsolved constraints 5 =? 1, 1 =? 5
\end{verbatim}
%
That is, LabMate says that for \texttt{C} to have a sensible type, we need the (impossible) constraints $5 = 1$ and $1 = 5$ to be true, arising from the fact that we have declared \texttt{B} to be a $5 \times 1$ matrix, but in fact it is a $1 \times 5$ matrix. Note that the original source code has been retained, but a response has been inserted to our \texttt{typeof} query. If we change \texttt{B} to
\begin{verbatim}
B = [ a; aCal; c; cCal; m ]
\end{verbatim}
instead, then LabMate responds
\begin{verbatim}
%< C :: [ 12 x 1 ] double
\end{verbatim}
This way, LabMate can detect matrix size errors at development time, rather than at runtime.

>>>>

\title{LabMate: supporting types for MATLAB}
\author{Conor McBride  \affiliationmark{1}, Georgi Nakov  \affiliationmark{1}, Fredrik Nordvall Forsberg  \affiliationmark{1}, Alistair Forbes  \affiliationmark{2}, Keith Lines  \affiliationmark{2}}
\affiliation{%
\affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , UK}  \\
\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, UK}}

@@

This calculation can be converted into a MATLAB problem with LabMate annotations as follows, using MATLAB's builtin left division operator \texttt{A \textbackslash{} b} to compute a least squares fit. We then ask for the type of the resulting solution. Note that for reasons of space only twelve values are listed. A real measurement would contain considerably more values:
\begin{verbatim}
%> A :: [ 12 x 5 ] double
A = [ 1  0  1  0  t1
      1  0  1  0  t2
      1  0  1  0  t3
     -1  0  1  0  t4
     -1  0  1  0  t5
     -1  0  1  0  t6
      0  1  0  1  t7
      0  1  0  1  t8
      0  1  0  1  t9
      0 -1  0  1  t10
      0 -1  0  1  t11
      0 -1  0  1  t12 ]

%> b :: [ 12 x 1 ] double
b = [ v1 v2 v3 v4 v5 v6 ...
        v7 v8 v9 v10 v11 v12 ]

x = A \ b

%> typeof x
\end{verbatim}
%
The lines starting with `\texttt{\%>}' are interpreted as ordinary comments by MATLAB, but as \remph{input directives} by LabMate. LabMate acts by reading the source code and outputting it again, but possibly with some additional \remph{responses} to the directives added.

Did you notice the error in the above code? The vector \texttt{b} should be of size $12 \times 1$, but is erroneously of size $1 \times 12$ instead. LabMate detects this error and produces the following as output.
\begin{verbatim}
%> A :: [ 12 x 5 ] double
A = [ 1  0  1  0  t1
      1  0  1  0  t2
      1  0  1  0  t3
     -1  0  1  0  t4
     -1  0  1  0  t5
     -1  0  1  0  t6
      0  1  0  1  t7
      0  1  0  1  t8
      0  1  0  1  t9
      0 -1  0  1  t10
      0 -1  0  1  t11
      0 -1  0  1  t12 ]

%> b :: [ 12 x 1 ] double
%< unsolved constraints 12 =? 1, 1 =? 12
b = [ v1 v2 v3 v4 v5 v6 ...
        v7 v8 v9 v10 v11 v12 ]

x = A \ B

%> typeof x
%< the type of x is quite a puzzle
\end{verbatim}
%
That is, LabMate says that for \texttt{x} to have a sensible type, we need the (impossible) constraints $5 = 1$ and $1 = 5$ to be true, arising from the fact that we have declared \texttt{b} to be a $12 \times 1$ matrix, but in fact it is a $1 \times 12$ matrix. Note that the original source code has been retained exactly as is, including whitespace characters, but a response has been inserted to our \texttt{typeof} query. If we change \texttt{b} to
\begin{verbatim}
b = [ v1 v2 v3 v4 v5 v6 ...
        v7 v8 v9 v10 v11 v12 ]'
\end{verbatim}
using MATLAB's transpose operator instead, then LabMate responds
\begin{verbatim}
%< x :: [ 5 x 1 ] double
\end{verbatim}
This way, LabMate can detect matrix size errors at development time, rather than at runtime --- running the original code would indeed crash with the error \textcolor{red}{\textbf{\texttt{Matrix dimensions must agree}}}.
